---
title: "Exploratory Analysis of Students Performance in Exams"
author: "Nicolette Johnson"
date: "October 9, 2019"
output:
  html_document:
    fig_width: 10
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's load the packages.
```{r}
# Install and load packages
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(gridExtra) # Get the gridExtra package for grid.arrange()
library(waffle)
```

Let's now import, clean, and check the data for missing values.
```{r}
# Import raw data
rawdata <- read.csv(file = "data/StudentsPerformance.csv", header = TRUE)
```

<br><br>

### Cleaning the data

According to the dataset's [page](https://www.kaggle.com/spscientist/students-performance-in-exams) on Kaggle, there is no indication of missing value codes (e.g., -9999 or anything similar).

I will just run a simple check to see if there are any missing values:
```{r}
# Checking for missing values:
cat("There are", sum(is.na(rawdata)), "missing values.")
```

Great! Our data contains zero missing values, so we can start looking into the data.

```{r}
# Get a feel for the data
str(rawdata)
head(rawdata, n = 10)
summary(rawdata)
```
Overall, the data is very neat and straightforward. Among the 8 variables, 5 are categorical and 3 are numerical.


Next, let's convert the raw data into a tibble data frame for easier data analysis. It would also be a good idea to convert certain catagorical data into ordinal data. 

### Data Manipulation
```{r}
# Converting raw data into a tibble
spdata <- as_tibble(rawdata)
# Converting appropriate categorical data to ordinal data
paredu <- ordered(spdata$parental.level.of.education, levels = c("some high school", "high school", "some college", "associate's degree", "bachelor's degree", "master's degree"))
```
Here, I have created a new variable called ***paredu*** which contains the parental level of education in an ordered format. This will come in handy later when we look at visualizations.

Let's look at the spread of students among different race/ethnicity groups:
```{r}
# Waffle chart for race/ethnicity:
race_counts <- spdata %>%
  group_by(race.ethnicity) %>%
    summarize(number_of_students = n()) %>%
    mutate(percent = round(number_of_students/sum(number_of_students)*100)) # This finds the percentage of the total for each race/enthnicity group
race_counts # Prints the race_counts table
percent_count <- race_counts$percent # Created an array with the new percent column
names(percent_count) <-race_counts$race.ethnicity # Named the percent array with the corresponding Group name
waffle(percent_count, title = "Spread of Students in Each Race/Ethnicity Group",
       xlab = "1 square = 1% of the total number of students")
```

With the waffle chart, we can see the spread of students in each race/ethnicity group. Here, each colored square represents 1% of students. For instance, Group A has 9 squares and thus represents 9% of our total students in the dataset.

<br>

#### *Grading Scale*
I am interpreting the grading scale as follows:

* A: 90-100
* B: 80-89
* C: 70-79
* D: 60-69
* F: 0-59

I will use this scale to place scores into "buckets" (i.e., grades).

```{r}
# Creating new grade columns based on corresponding scores:
spdata_with_grades <- spdata %>%
  mutate(math.grade = case_when(math.score < 60 ~ "F",
                                math.score >= 60 & math.score <= 69 ~ "D",
                                math.score >= 70 & math.score <= 79 ~ "C",
                                math.score >= 80 & math.score <= 89 ~ "B",
                                math.score >= 90 & math.score <= 100 ~ "A"),
         reading.grade = case_when(reading.score < 60 ~ "F",
                                   reading.score >= 60 & reading.score <= 69 ~ "D",
                                   reading.score >= 70 & reading.score <= 79 ~ "C",
                                   reading.score >= 80 & reading.score <= 89 ~ "B",
                                   reading.score >= 90 & reading.score <= 100 ~ "A"),
         writing.grade = case_when(writing.score < 60 ~ "F",
                                   writing.score >= 60 & writing.score <= 69 ~ "D",
                                   writing.score >= 70 & writing.score <= 79 ~ "C",
                                   writing.score >= 80 & writing.score <= 89 ~ "B",
                                   writing.score >= 90 & writing.score <= 100 ~ "A"))
```

Let's check to see if my newly-created buckets worked:

```{r}
# Checking to see if each new grade has appropriate scores:
math_check <- spdata_with_grades %>%
  ggplot(aes(x = math.grade, y = math.score)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
reading_check <- spdata_with_grades %>%
  ggplot(aes(x = reading.grade, y = reading.score)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
writing_check <- spdata_with_grades %>%
  ggplot(aes(x = writing.grade, y = writing.score)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
grid.arrange(math_check, reading_check, writing_check, ncol = 3)
```

Great, it looks like the buckets catch the appropriate values! It's also good to see the outliers in the data (very low grades, < 30). Later, we can see if these scores point to the same students or not. Some students may have low grades in all 3 subjects, or they may have higher grades in one or two subjects but struggle in others. The plan would be to help these struggling students achieve better scores.

And just to be safe, let's check for missing data again since I created new buckets.

```{r}
# Checking for missing values again. This time, using the newly-created dataframe:
cat("There are", sum(is.na(spdata_with_grades)), "missing values.")
```

It looks there are still zero missing values!

Let's look at our data with the newly-added grade columns.
```{r}
# Let's look at our data with the new columns:
str(spdata_with_grades)
select(spdata_with_grades, math.grade:writing.grade)
```

We can see that the new columns are listed as characters.

Let's convert them to factors with levels using lapply:
```{r}
# The new columns need to be converted to factors:
grades <- c("math.grade", "reading.grade", "writing.grade")
spdata_with_grades[grades] <- lapply(spdata_with_grades[grades], factor)
str(spdata_with_grades)
```
This will also be useful later when we get into visualization.

Before we move on, let's write the new data into a new file. This will keep our original data intact.
```{r}
# Writing to a new file:
write.csv(spdata_with_grades, file = "data/StudentsPerformance-wrangled.csv", row.names = FALSE, col.names = TRUE)
```
<br><br>

### Data Visualization

```{r}
# Let's look at math scores
math_scores_dist <- ggplot(spdata_with_grades, aes(x = math.score)) +
  geom_histogram() +
  ggtitle("Math Scores Distribution") +
  xlab("Math Score") +
  coord_cartesian(ylim = c(0, 120))
# Let's look at reading scores
read_scores_dist <- ggplot(spdata_with_grades, aes(x = reading.score)) +
  geom_histogram() +
  ggtitle("Reading Scores Distribution") +
  xlab("Reading Score") +
  coord_cartesian(ylim = c(0, 120))
# Let's look at writing scores
writ_scores_dist <- ggplot(spdata_with_grades, aes(x = writing.score)) +
  geom_histogram() +
  ggtitle("Writing Scores Distribution") +
  xlab("Writing Score") +
  coord_cartesian(ylim = c(0, 120))
grid.arrange(math_scores_dist, read_scores_dist, writ_scores_dist, ncol = 3)
```

We can see the math, reading, and writing scores all follow a fairly normal distribution with some outliers (left-skewed).

```{r fig.height = 5, fig.width = 10}
# Better Legend:
better_legend <- theme(
    legend.position = c(.99, .99),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(0, 0, 0, 0, "cm"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.key.size = unit(0.4, "cm")
    )
# Long Title Wrap function:
wrapper <- function(x, ...)
{
  paste(strwrap(x, ...), collapse = "\n")
}
# Here are the math grades vs. parental level of education
math_grades_vs_paredu <- ggplot(spdata_with_grades, aes(x = paredu, fill = math.grade)) +
  geom_bar(position = "dodge") +
  ggtitle(wrapper("Math Grades Grouped by Parental Level of Education", width = 30)) +
  xlab("Parental Level of Education") +
  labs(fill = "Math Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
# Here are the reading grades vs. parental level of education
read_grades_vs_paredu <- ggplot(spdata_with_grades, aes(x = paredu , fill = reading.grade)) +
  geom_bar(position = "dodge") +
   ggtitle(wrapper("Reading Grades Grouped by Parental Level of Education", width = 30)) +
  xlab("Parental Level of Education") +
  labs(fill = "Reading Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
# Here are the writing grades vs. parental level of education
writ_grades_vs_paredu <- ggplot(spdata_with_grades, aes(x = paredu, fill = writing.grade)) +
  geom_bar(position = "dodge") +
  ggtitle(wrapper("Writing Grades Grouped by Parental Level of Education", width = 30)) +
  xlab("Parental Level of Education") +
  labs(fill = "Writing Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
grid.arrange(math_grades_vs_paredu, read_grades_vs_paredu, writ_grades_vs_paredu, ncol = 3)
```

This gives us general idea of how the grades look for each type of parental education.

However, to get a better understanding, let's look at a proportional graph of the same data:

```{r fig.height = 5, fig.width = 13}
# Proportional graph of math grades vs. parental level of education
math_grades_vs_paredu_prop <-ggplot(spdata_with_grades, aes(x = paredu, fill = math.grade)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Math Grades Grouped by Parental Level of Education", width = 40)) +
  xlab("Parental Level of Education") +
  ylab("Proportion") +
  labs(fill = "Math Grade") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of reading grades vs. parental level of education
read_grades_vs_paredu_prop <- ggplot(spdata_with_grades, aes(x = paredu, fill = reading.grade)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Reading Grades Grouped by Parental Level of Education", width = 40)) +
  xlab("Parental Level of Education") +
  ylab("Proportion") +
  labs(fill = "Reading Grade") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of writing grades vs. parental level of education
writ_grades_vs_paredu_prop <- ggplot(spdata_with_grades, aes(x = paredu, fill = writing.grade)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Writing Grades Grouped by Parental Level of Education", width = 40)) +
  xlab("Parental Level of Education") +
  ylab("Proportion") +
  labs(fill = "Writing Grade") +
  theme(axis.text.x = element_text(angle = 90))
grid.arrange(math_grades_vs_paredu_prop, read_grades_vs_paredu_prop, writ_grades_vs_paredu_prop, ncol = 3)
```

From this view, we can see that the higher three levels of parental education (master's degree, bachelor's degree, and associate's degree) tend to have a higher proportion of students with As, Bs, and Cs compared to the lower three levels (some college, high school, and some high school).

And here is a proportional graph of grades vs. parental level of education:

```{r fig.height = 4, fig.width = 15}
# Proportional graph of math grades vs. parental level of education
math_grades_vs_paredu_prop_flip <- ggplot(spdata_with_grades, aes(x = math.grade, fill = paredu)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Parental Level of Education Grouped by Math Grades", width = 40)) +
  xlab("Math Grades") +
  ylab("Proportion") +
  labs(fill = "Parental Education") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of reading grades vs. parental level of education
read_grades_vs_paredu_prop_flip <- ggplot(spdata_with_grades, aes(x = reading.grade, fill = paredu)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Parental Level of Education Grouped by Reading Grades", width = 40)) +
  xlab("Reading Grades") +
  ylab("Proportion") +
  labs(fill = "Parental Education") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of writing grades vs. parental level of education
writ_grades_vs_paredu_prop_flip <- ggplot(spdata_with_grades, aes(x = writing.grade, fill = paredu)) +
  geom_bar(position = "fill") +
  ggtitle(wrapper("Proportion of Parental Level of Education Grouped by Writing Grades", width = 40)) +
  xlab("Writing Grades") +
  ylab("Proportion") +
  labs(fill = "Parental Education") +
  theme(axis.text.x = element_text(angle = 90))
grid.arrange(math_grades_vs_paredu_prop_flip, read_grades_vs_paredu_prop_flip, writ_grades_vs_paredu_prop_flip, ncol = 3)
```

When we switch the view, we can easily see the proportion of each type of parental level of education for each grade.


Now, let's look at all grades vs. the race/ethnicity of the student.

```{r fig.height = 5, fig.width = 12}
# Here are the math grades vs. race/ethnicity
math_grades_vs_race <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = math.grade)) +
  geom_bar(position = "dodge") +
  ggtitle("Math Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  labs(fill = "Math Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
# Here are the reading grades vs. race/ethnicity
read_grades_vs_race <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = reading.grade)) +
  geom_bar(position = "dodge") +
  ggtitle("Reading Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  labs(fill = "Reading Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
# Here are the writing grades vs. race/ethnicity
writ_grades_vs_race <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = writing.grade)) +
  geom_bar(position = "dodge") +
  ggtitle("Writing Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  labs(fill = "Writing Grade") +
  theme(axis.text.x = element_text(angle = 90)) +
  better_legend
grid.arrange(math_grades_vs_race, read_grades_vs_race, writ_grades_vs_race, ncol = 3)
```

Like before, we can get more insight if we view this as a proportional graph:

```{r fig.height = 5, fig.width = 15}
# Proportional graph of math grades vs. race/ethnicity
math_grades_vs_race_prop <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = math.grade)) +
  geom_bar(position = "fill") +
  ggtitle("Math Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  ylab("Proportion") +
  labs(fill = "Math Grade") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of reading grades vs. race/ethnicity
read_grades_vs_race_prop <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = reading.grade)) +
  geom_bar(position = "fill") +
  ggtitle("Reading Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  ylab("Proportion") +
  labs(fill = "Reading Grade") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of writing grades vs. race/ethnicity
writ_grades_vs_race_prop <- ggplot(spdata_with_grades, aes(x = race.ethnicity, fill = writing.grade)) +
  geom_bar(position = "fill") +
  ggtitle("Writing Grades Grouped by Race/Ethnicity") +
  xlab("Race/Ethnicity") +
  ylab("Proportion") +
  labs(fill = "Writing Grade") +
  theme(axis.text.x = element_text(angle = 90))
grid.arrange(math_grades_vs_race_prop, read_grades_vs_race_prop, writ_grades_vs_race_prop, ncol = 3)
```

We can see that Group E received the highest proportion of As and Bs compared to other groups in all 3 subjects. It seems as if Group A seems to struggle more in all 3 subjects since the largest proportion of grades are Fs for this group.

```{r fig.height = 5, fig.width = 15}
# Proportional graph of race/ethnicity vs. math grades
math_grades_vs_race_prop_flip <- ggplot(spdata_with_grades, aes(x = math.grade, fill = race.ethnicity)) +
  geom_bar(position = "fill") +
  ggtitle("Proportion of Race/Ethnicity Grouped by Math Grades") +
  xlab("Math Grades") +
  ylab("Proportion") +
  labs(fill = "Race/Ethnicity") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of race/ethnicity vs. reading grades
read_grades_vs_race_prop_flip <- ggplot(spdata_with_grades, aes(x = reading.grade, fill = race.ethnicity)) +
  geom_bar(position = "fill") +
  ggtitle("Proportion of Race/Ethnicity Grouped by Reading Grades") +
  xlab("Reading Grades") +
  ylab("Proportion") +
  labs(fill = "Race/Ethnicity") +
  theme(axis.text.x = element_text(angle = 90))
# Proportional graph of race/ethnicity vs. writing grades
writ_grades_vs_race_prop_flip <- ggplot(spdata_with_grades, aes(x = writing.grade, fill = race.ethnicity)) +
  geom_bar(position = "fill") +
  ggtitle("Proportion of Race/Ethnicity Grouped by Writing Grades") +
  xlab("Writing Grades") +
  ylab("Proportion") +
  labs(fill = "Race/Ethnicity") +
  theme(axis.text.x = element_text(angle = 90))
grid.arrange(math_grades_vs_race_prop_flip, read_grades_vs_race_prop_flip, writ_grades_vs_race_prop_flip, ncol = 3)
```

Here, we can see that of all grades that result in an F, Group C makes up the highest proportion. Whereas for all grades that result in A, Group E makes up the largest proportion.

We can also display these in a faceted barchart in order to compare the distribution of grades among different groups:
```{r}
ggplot(spdata_with_grades, aes(x = math.grade)) +
  geom_bar() +
  facet_wrap(~race.ethnicity, ncol = 5) +
  ggtitle("Math Grades Distribution by Race/Ethnicity") +
  xlab("Math Grades")
```

Overall, math grades are left-skewed for Groups A, B, and C meaning there are much fewer As than Fs. Group D exhibits a left tail that is far less extreme than the first 3 groups. And Group E seems to have an almost perfectly uniform distribution of grades within the group.

```{r}
ggplot(spdata_with_grades, aes(x = reading.grade)) +
  geom_bar() +
  facet_wrap(~race.ethnicity, ncol = 5) +
  ggtitle("Reading Grades Distribution by Race/Ethnicity") +
  xlab("Reading Grades")
```

Reading grades are left-skewed for Groups A and B. This time, Group C shows a more normal distribution, but is still left-skewed. Group D also exhibits a somewhat normal distribtuion but again is still left-skewed. And Group E also shows a slightly normal distribution for reading grades.

```{r}
ggplot(spdata_with_grades, aes(x = writing.grade)) +
  geom_bar() +
  facet_wrap(~race.ethnicity, ncol = 5) +
  ggtitle("Writing Grades Distribution by Race/Ethnicity") +
  xlab("Writing Grades")
```

Writing grades are left-skewed for Groups A, B, and C. Group D shows a somewhat normal distribtuion but again is still left-skewed. And Group E also shows a slightly normal distribution for writing grades.


Remember that in our proportion graphs, Groups A and B seemed to struggle the most as indicated with the most Fs compared to the other groups.

And remember our boxplot of our scores from earlier? Here are just the failing scores:

```{r fig.height = 5, fig.width = 10}
# Subset the failing scores:
math_Fs <- spdata_with_grades %>%
  filter(math.grade == "F") %>%
  ggplot(aes(x = math.grade, y = math.score)) +
  geom_boxplot() +
  ggtitle("Math Scores Below 60") +
  xlab("Math Grade") +
  ylab("Math Score") +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
reading_Fs <- spdata_with_grades %>%
  filter(reading.grade == "F") %>%
  ggplot(aes(x = reading.grade, y = reading.score)) +
  geom_boxplot()  +
  ggtitle("Reading Scores Below 60") +
  xlab("Reading Grade") +
  ylab("Reading Score") +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
writing_Fs <- spdata_with_grades %>%
  filter(writing.grade == "F") %>%
  ggplot(aes(x = writing.grade, y = writing.score)) +
  geom_boxplot() +
  ggtitle("Writing Scores Below 60") +
  xlab("Writing Grade") +
  ylab("Writing Score") +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 5))
grid.arrange(math_Fs, reading_Fs, writing_Fs, ncol = 3)
```

The outliers seem to start at scores of 30 and end at 0. The plan is to find out more information about the students behind these scores, (e.g., find out if the student is struggling in all 3 areas or just 1 or 2, find out their demographics and socioeconomic status, etc.)

Let's first look at the data on these students:
```{r fig.height = 7, fig.width = 10}
# Create a Student ID field for a unique identifier:
spdata_with_grades_ID <- tibble::rowid_to_column(spdata_with_grades, "ID")

# Convert new ID variable to factor:
spdata_with_grades_ID$ID <- as.factor(spdata_with_grades_ID$ID)

# Filter for the <= 30 score students that need help:
math_below30 <- spdata_with_grades_ID %>%
  filter(math.score <= 30)
reading_below30 <- spdata_with_grades_ID %>%
  filter(reading.score <= 30)
writing_below30 <- spdata_with_grades_ID %>%
  filter(writing.score <= 30)

# Create a union for these 3 sets of data:
students_below30 <- list(math_below30, reading_below30, writing_below30) %>%
  reduce(union, by = "ID")

# Visualize the data:
students_below30.long <- gather(students_below30, key = "Subject", value = "Score", -ID, -gender, -race.ethnicity, -parental.level.of.education, -lunch, -test.preparation.course, -math.grade, -reading.grade, -writing.grade)
ggplot(students_below30.long, aes(x = ID, y = Score, fill = Subject)) +
  geom_col(position = "dodge", color = "black", width = 0.65) +
  geom_hline(yintercept = 30, linetype = "dotted") +
  ggtitle("Students with Scores Below 30") +
  xlab("Student ID") +
  theme_bw() +
  coord_flip()
```

There are 18 students that have a score of 30 or below in at least 1 subject. As we can see, there are some students who score 30 or below in all 3 subjects and some who score higher in one or two other subjects. One student, #60, scored 0 in math and also has the lowest reading and writing scores out of any other student.

Let's look at the characteristics of these struggling students next.

```{r}
# Look at race/ethnicity:
# Dummy variable for race/ethnicity:
students_below30 <- students_below30 %>%
  mutate(Groups = "Groups")
students_below30_race_prop <- ggplot(students_below30, aes(x = Groups, fill = race.ethnicity)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of Students Who Scored Below 30 by Race/Ethnicity", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Race/Ethnicity")
# Look at parental level of education:
# Dummy variable for parental level of education:
students_below30 <- students_below30 %>%
  mutate(ParentalEducation = "Parental Education")
# Create new paredu for updated dataset so data will be in order:
paredu_below30 <- ordered(students_below30$parental.level.of.education, levels = c("some high school", "high school", "some college", "associate's degree", "bachelor's degree", "master's degree"))
students_below30_paredu_prop <- ggplot(students_below30, aes(x = ParentalEducation, fill = paredu_below30)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of Students Who Scored Below 30 by Parental Education", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Parental Education")
# Look at lunch type (proxy for socioeconomic status):
# Dummy variable for lunch:
students_below30 <- students_below30 %>%
  mutate(Lunch = "Lunch Type")
students_below30_lunch_prop <- ggplot(students_below30, aes(x = Lunch, fill = lunch)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of Students Who Scored Below 30 by Lunch Type", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Lunch Type")
# Look at test prep:
# Dummy variable for test prep:
students_below30 <- students_below30 %>%
  mutate(TestPrep = "Test Prep Course")
students_below30_testprep_prop <- ggplot(students_below30, aes(x = TestPrep, fill = test.preparation.course)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of Students Who Scored Below 30 by Test Prep Course", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Test Prep Course")
grid.arrange(students_below30_race_prop, students_below30_paredu_prop, students_below30_lunch_prop, students_below30_testprep_prop, ncol = 4)
```

From these graphs, we can gain information about these struggling students.

A large proportion of the students come from race/ethnicity groups B and C.

Roughly 65% of these students have parents who did not attend college.

In this project, I am using lunch type as a proxy for socioeconomic status. I am assuming that if a student receives free/reduced lunch, they are of low socioeconomic status. Around 85% of these students receive free/reduced lunch and thus are of low socioeconomic status.

And finally, about 88% of these students have not completed a test preparation course.

<br><br>

Let's compare these struggling students to the students who get the highest grades:

```{r}
# Create a new combined score column:
spdata_with_grades_ID <- spdata_with_grades_ID %>% mutate(combined.score = math.score + reading.score + writing.score)

# Let's look at the top 18 students since we had 18 in the struggling student group:
spdata_with_grades_ID %>% top_n(18, combined.score)
```

The code brings back 19 students. There is likely a tie in the combined score variable. Let's try to filter for the scores that are 90 and above:

```{r}
# Filter for the scores that are >= 90:
students_top18 <- spdata_with_grades_ID %>%
  top_n(18, combined.score) %>%
  filter(math.score >= 90 & reading.score >= 90 & writing.score >= 90)
students_top18
```

It looks like that worked and brought us back the top 18 students! Let's celebrate by making more visualizations!

```{r}
# Look at race/ethnicity:
# Dummy variable for race/ethnicity:
students_top18 <- students_top18 %>%
  mutate(Groups = "Groups")
students_top18_race_prop <- ggplot(students_top18, aes(x = Groups, fill = race.ethnicity)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of The Top 18 Students by Race/Ethnicity", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Race/Ethnicity")
# Look at parental level of education:
# Dummy variable for parental level of education:
students_top18 <- students_top18 %>%
  mutate(ParentalEducation = "Parental Education")
# Create new paredu for updated dataset so data will be in order:
paredu_top18 <- ordered(students_top18$parental.level.of.education, levels = c("some high school", "high school", "some college", "associate's degree", "bachelor's degree", "master's degree"))
students_top18_paredu_prop <- ggplot(students_top18, aes(x = ParentalEducation, fill = paredu_top18)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of The Top 18 Students by Parental Education", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Parental Education")
# Look at lunch type (proxy for socioeconomic status):
# Dummy variable for lunch:
students_top18 <- students_top18 %>%
  mutate(Lunch = "Lunch Type")
students_top18_lunch_prop <- ggplot(students_top18, aes(x = Lunch, fill = lunch)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of The Top 18 Students by Lunch Type", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Lunch Type")
# Look at test prep:
# Dummy variable for test prep:
students_top18 <- students_top18 %>%
  mutate(TestPrep = "Test Prep Course")
students_top18_testprep_prop <- ggplot(students_top18, aes(x = TestPrep, fill = test.preparation.course)) +
  geom_bar(position = "fill", width = 0.5) +
  ggtitle(wrapper("Proportion of The Top 18 Students by Test Prep Course", width = 20)) +
  xlab(NULL) +
  ylab("Proportion") +
  labs(fill = "Test Prep Course")
grid.arrange(students_top18_race_prop, students_top18_paredu_prop, students_top18_lunch_prop, students_top18_testprep_prop, ncol = 4)
```

From these graphs, we can gain information about the top 18 students.

A large proportion of the students come from race/ethnicity groups D and E.

Roughly 56% of these students have parents received either bachelor's or master's degrees.

In this project, I am using lunch type as a proxy for socioeconomic status. I am assuming that if a student receives free/reduced lunch, they are of low socioeconomic status. Only 16% of these students receive free/reduced lunch and thus are of low socioeconomic status.

And finally, about 33% of these students have not completed a test preparation course.

<br><br>

Let's compare these two groups of students to each other.

```{r}

```

